Contracts for unions and intersections are hard for Nickel

### 2021-05-27 paper.tex
read on intro

However, static text-based configurations show their limits beyond a certain
size and complexity. Data cannot be transformed, combined, shared, nor validated
(without resorting to separate schemas and tools).  This lack of expressivity
results at best in boilerplate and duplication of information, but it can also
make data invalid or inconsistent.  Correctly configuring a modern system is ard
and failures may have substantial negative consequences.

@@

Assertions are usually mainly used for correctness by enforcing pre- and
post-conditions of functions at runtime. In their foundational
paper\cite{FindlerFelleisenHOContracts}, Findler and Felleisen introduce a
principled approach to run-time assertion checking called contracts. As opposed
to plain assertions, contracts can be composed naturally into more complex ones.
Nickel features such a contract system. For example, contracts nicely supports
higher-order functions. If the user has written an \nickel{Even} contract for
number, they can use it right away to form a \nickel{#Even -> #Int} contract, as
illustrated in Figure~\ref{fig:contract-higher-order}. This contracts will check
at each call that the argument is an even number (pre-condition) and that the
return value is a integer (post-condition). The \nickel{#} character
distinguishes contracts in Nickel. Contracts also introduce the notion of blame
which is crucial to good error reporting in presence of higher-order contracts
and other composed contracts.

@@

In our opinion, one of the main appeal of Nickel is this ability to write a
simple validation function for, say, a port number, and have it immediately
available as a contract that can be used to construct a other contracts by
writing natural type-like expressions. Another example is given in
Figure~\ref{fig:contract-composition}. The first line defines a port contract
from a predicate and use it in the definition of a larger
\lstinline+Configuration+ record contract. The \lstinline+Configuration+
contract checks that a value is a record with the following shape: it must have
a \lstinline+port+ field obeying the \lstinline+Port+ contract, and a
\lstinline+host+ field obeying a \lstinline+Host+ contract. It plays the role of
a \emph{schema} used to validate the final data, which is exactly what we were
looking for in the previous section. Not only it plays the role of a schema, but
it also \emph{looks} like one, as opposed to a less readable plain validation
function. The \lstinline+|+ operator entails that a value respect a contract,
akin to \lstinline+:+ for types. We assume that a \lstinline+Host+ contract has
been defined in a similar way. \unsure{What's missing in this section is an
example of how contracts differ from standard dynamic type checking on some
function. Namely, the quality of error messages.}

\begin{figure}
  \begin{center}
\begin{lstlisting}[language=nickel]
let Port = contracts.fromPred (fun value =>
  builtins.isNum value && value % 1 == 0
  && value >= 0 && value <= 65353) in
let Configuration = {
  host | #Host,
  port | #Port,
} in
{host = "nickel-lang.org", port = 80}
| #Configuration
\end{lstlisting}
  \end{center}
\caption{Contract composition}
\label{fig:contract-composition}
\end{figure}

Contracts have another perk: they correspond to the type casts of gradual
typing, a discipline for mixing static and dynamic typing in the same language.
As such, they are a precious ingredient for handling the interaction of
statically typed and dynamically typed code. While we focused on the writing of
configuration, Nickel is a proper programming language that is also designed to
write generic and reusable library code. There, the typing trade-off is
different, and static typing is more adapted. Nickel has a gradual type system,
which is yet another argument for having contracts. Let us end this section by
stressing the following points:
\begin{itemize}
    \item \textbf{Even in a purely dynamically typed language, contracts are an
        appealing feature for validation and correctness}. The gradual
        typing part deserved to be mentioned, but is is out of the scope of this
        paper. 
    \item \textbf{User-defined contracts are essential for validation}. Some contract
        systems (as the one of~\ref{RootCauseOfBlame} that will be extensively
        discussed later in the paper) do not feature user-defined contracts.
        Rather, a fixed set of built-in contracts such as \nickel{Number} or
        \nickel{String} are provided, very much like base types. The absence of
        user-defined contracts is understandable when using contracts as a
        dynamic typing discipline for correctness. But a system without
        user-defined contracts is limited in the same way as static typing with
        respect to validation.
\end{itemize}

>>>>

However, static text-based configurations show their limits beyond a certain
size and complexity. Data cannot be transformed, combined, shared, nor validated
(without resorting to separate schemas and tools).  This lack of expressivity
results at best in boilerplate and duplication of information, but it can also
make data invalid or inconsistent.  Correctly configuring a modern system is hard
and failures may have substantial negative consequences.

@@

Assertions are usually mainly used for correctness by enforcing pre- and
post-conditions of functions at runtime. In their foundational
paper\cite{FindlerFelleisenHOContracts}, Findler and Felleisen introduce a
principled approach to run-time assertion checking called contracts. As opposed
to plain assertions, contracts can be composed naturally into more complex ones.
Nickel features such a contract system. For example, contracts nicely support
higher-order functions. If the user has written an \nickel{Even} contract for
number, they can use it right away to form a \nickel{#Even -> #Int} contract, as
illustrated in Figure~\ref{fig:contract-higher-order}. This contract will check
at each call that the argument is an even number (pre-condition) and that the
return value is a integer (post-condition). The \nickel{#} character
distinguishes contracts in Nickel. Contracts also introduce the notion of blame
which is crucial to good error reporting in presence of higher-order contracts
and other composed contracts.

@@

In our opinion, one of the main appeals of Nickel is this ability to write a
simple validation function for, say, a port number, and have it immediately
available as a contract that can be used to construct other contracts by
writing natural type-like expressions. Another example is given in
Figure~\ref{fig:contract-composition}. The first line defines a port contract
from a predicate and use it in the definition of a larger
\lstinline+Configuration+ record contract. The \lstinline+Configuration+
contract checks that a value is a record with the following shape: it must have
a \lstinline+port+ field obeying the \lstinline+Port+ contract, and a
\lstinline+host+ field obeying a \lstinline+Host+ contract. It plays the role of
a \emph{schema} used to validate the final data, which is exactly what we were
looking for in the previous section. Not only it plays the role of a schema, but
it also \emph{looks} like one, as opposed to a less readable plain validation
function. The \lstinline+|+ operator entails that a value respects a contract,
akin to \lstinline+:+ for types. We assume that a \lstinline+Host+ contract has
been defined in a similar way. \unsure{What's missing in this section is an
example of how contracts differ from standard dynamic type checking on some
function. Namely, the quality of error messages.}

\begin{figure}
  \begin{center}
\begin{lstlisting}[language=nickel]
let Port = contracts.fromPred (fun value =>
  builtins.isNum value && value % 1 == 0
  && value >= 0 && value <= 65353) in
let Configuration = {
  host | #Host,
  port | #Port,
} in
{host = "nickel-lang.org", port = 80}
| #Configuration
\end{lstlisting}
  \end{center}
\caption{Contract composition}
\label{fig:contract-composition}
\end{figure}
\unsure{I wouldn't use nickel-lang.org unless you are
planning on putting something up there. Maybe nixos.org?}

Contracts have another perk: they correspond to the type casts of gradual
typing, a discipline for mixing static and dynamic typing in the same language.
As such, they are a precious ingredient for handling the interaction of
statically typed and dynamically typed code. While we focused on the writing of
configuration, Nickel is a proper programming language that is also designed to
write generic and reusable library code. There, the typing trade-off is
different, and static typing is more adapted. Nickel has a gradual type system,
which is yet another argument for having contracts. Let us end this section by
stressing the following points:
\begin{itemize}
    \item \textbf{Even in a purely dynamically typed language, contracts are an
        appealing feature for validation and correctness}. The gradual
        typing part deserved to be mentioned, but is is out of the scope of this
        paper. 
    \item \textbf{User-defined contracts are essential for validation}. Some contract
        systems (as the one of~\cite{RootCauseOfBlame} that will be extensively
        discussed later in the paper) do not feature user-defined contracts.
        Rather, a fixed set of built-in contracts such as \nickel{Number} or
        \nickel{String} are provided, very much like base types. The absence of
        user-defined contracts is understandable when using contracts as a
        dynamic typing discipline for correctness. But a system without
        user-defined contracts is limited in the same way as static typing with
        respect to validation.
\end{itemize}
